Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-07 kappa.tex
Small corrections in field area description.

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes. The Laurentide ice
sheet, largest of the three ice sheets, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, our region of focus in the present thesis.

@@

In North American, the Western Cordillera forms a broad mountain system with an
average east-west width of a about 1000\,km. Its topography is a complex
arrangement of mountain ranges and intramontane basins. It was formed during
multiple accretion phases associated with the closure of several back-arc
oceanic trenches over a 200\,Ma-long geologic history.

Most of the major reliefs are found along two major bands that run parallel to
each other along the general orientation of the Cordillera. The western band
consists of the Pacific Coast orogenic belt. Major ice-covered reliefs include,
from North to South, the Alaska Range, the Wranglell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern band consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
rectilinear Rocky Mountain / Tintina Trench. It includes the Selwyn and Mac
Kenzie Mountains to the north, and the Rocky Mountains to the South. The brooks
range, in northern Alaska, has been glaciated but its glaciers were not
connected to the main body of the Cordilleran ice sheet.

The Cordillera comports several intermontane bansins, which, as we will see
further, host particular climate properties and significantly impacted patterns
of ice sheet flow. The two major depressions are the Interior Plateau of
central British Columbia, and the Liard Lowland. Between these two basins is
found a mildly elevated, but extensive and deeply incised mountain group known
as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past glacier dynamics in the Cordillera. Because of
the irregular arrangement of valleys and ridges that forms the ice sheet’s
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them. In terms of
numerical glacier modelling, it implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large, computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation over the model domain.
Due to the large latitudinal extent of the region, mean annual temperatures
vary between -12 and 12\,\degC. About half the model domain has a mean annual
temperature below freezing. However, seasonality, here computed as the
temperature difference between the warmest and the coldest months, varies
between around 10\,\degC along most of the Pacific Coast to above 40\,\degC in
the polar interior. This concerns most of the area above 60\degree~N, and
high-elevated mountain ranges below this latitude. There exist a partial
correlation between cold regions and those that have a large seasonality, so
that many regions where mean annual temperatures are well below freezing point
do experience warm summers.

>>>>

During the last glacial cycle, much of North America has been covered by ice.
At the last glacial maximum, apex of glaciation, three major ice sheets had
collided to form a continuous blanket of ice that spanned from Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago and the Great Lakes
\cref{fig:map-northamerica}. The Laurentide ice
sheet, largest of the three, was centred on the modern Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the Cordilleran ice sheet capped the mountain ranges of
the Western Cordillera, which forms our region of focus in the present thesis.

@@

In North American, the Western Cordillera forms a broad mountain system with an
average east-west width of about 1000\,km. Its topography is a complex
arrangement of mountain ranges and interior basins. It was formed during
multiple accretion phases associated with the closure of several back-arc
oceanic trenches over a 200\,Ma-long geological history \mref.

Most of the dominant reliefs are found along two major strips that run parallel
to each other along the general orientation of the Cordillera
\cref{fig:map-cordillera}. The western strip consists of the Pacific Coast
orogenic belt \mref. Major ice-covered reliefs include,
from North to South, the Alaska Range, the Wrangell and Saint-Elias Mountains,
the Coast Mountains and the North Cascades. The eastern strip consists of the
Nevadan and Laramide belts, which are separated by the deep and strikingly
linear Rocky Mountain / Tintina Trench \mref. It includes the Selwyn and
MacKenzie Mountains in the north, and the Rocky Mountains in the South. The
Brooks range, in northern Alaska, has been covered by ice, but its glaciers
were not connected to the main body of the Cordilleran ice sheet.

The Cordillera comports several intermontane basins that impacted patterns
of ice sheet flow. The two major depressions are the Interior Plateau of
central British Columbia, and the Liard Lowland. Between these two basins is
found a mildly elevated, but extensive and deeply incised mountain group known
as the Skeena Mountains.

The ubiquitously mountainous topography of the region presents a serious
obstacle to reconstructing past dynamics of the Cordilleran ice sheet. Because
of the irregular arrangement of valleys and ridges that formed the ice sheet's
floor, ice moved along tortuous and intertwined pathways, sometimes following
the major topographical troughs, and sometimes flowing across them. In terms of
numerical glacier modelling, it implies complex patterns of flow and stress
within the ice, and requires high-resolution, and thus, large computational
grids.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation over the model domain.
Due to the large latitudinal extent of the region, mean annual temperatures
vary between -12 and 12\,\degC. About half the model domain has a mean annual
temperature below freezing. However, seasonality, here computed as the
temperature difference between the warmest and the coldest months, varies
between around 10\,\degC\ along most of the Pacific Coast to above 40\,\degC\
in the polar interior. This concerns most of the area above 60\degree~N, and
high-elevated mountain ranges below this latitude. There exist a partial
correlation between cold regions and those that have a large seasonality, so
that many regions where mean annual temperatures are well below freezing point
do experience warm summers.
